---
title: "Untitled"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Exercise 3. Analysis of the student performance including alcohol consumption
## Introduction

There are two datasets available for performance in two different subjects: mathematics (mat) and Portuguese language (por).The database was created using two types of data: school reports and surveys, with the latter being used to supplement the preceding data. In this excercise we use a joined dataset, that has been combined by averaging (including the grade variables). 

To read the data in.

```{r}
alc <- read.csv("https://github.com/rsund/IODS-project/raw/master/data/alc.csv", sep = "," ,header=TRUE)
#print variable names
names(alc)

```

Student grades, demographic, socioeconomic, and school-related characteristics are among the data variables.
Alcohol consumption is given special consideration, as is its potential impact on school performance.
The daily and weekly alcohol consumption values are combined and averaged to create the variable alc use.
The desired variable is alcohol consumption, which is a binary (low/high) variable. 


##Four interesting variables to explain the alcohol consumption

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
