# 6. Analysis of longitudinal data

The examples from Vehkalahti and Everitt's (2019) book Multivariate Analysis for The Behavioral Sciences will be repeated, however the data sets will be swapped between the chapters. Each male subject was scored on the brief psychiatric rating scale (BPRS) before therapy began (week 0) and then at weekly intervals for eight weeks. The BPRS rates the severity of 18 symptom components, including aggression, suspiciousness, hallucinations, and grandiosity, on a scale of one to seven (extremely severe). The scale is used to assess patients who may be suffering from schizophrenia. 

But first we are going to focus on the RATS dataset. The BPRS fataset will be described later. The RATS dataset is describes as followed in the book: "we shall use some data from a nutrition study conducted in three groups of rats
(Crowder and Hand, 1990). The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period"

```{r}

library(dplyr)
library(ggplot2)

RATS <- read.csv("data/rats.csv")
RATS$ID <- as.factor(RATS$ID)
RATS$Group <- as.factor(RATS$Group)
str(RATS)
summary(RATS)

```

We can see that there are 16 rats (the highest number of ID is 16 as each rat has a own ID number). 

```{r}
ggplot(RATS, aes(x = time, y = weight, linetype= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, 2)) +
  facet_grid(. ~ Group, labeller=label_both) +
  theme(legend.position = "none") +
  scale_y_continuous(limits = c(min(RATS$weight), max(RATS$weight)))
```

The rats in Group 1 appear to be the tiniest. We can observe in group 2 that there is a one giant rat. The group of three rats appears to be larger as well. 

## The Golden Standardise

It is possible that rats that have high weights at the beginning of the study tend to have higher weights throughout the study.
```{r}
rats_tracked <- RATS %>% 
  group_by(time) %>%
  mutate(std.weight = (weight - mean(weight))/sd(weight)) %>%
  ungroup()


ggplot(rats_tracked, aes(x = time, y = std.weight, linetype= ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:8, 2)) +
  facet_grid(. ~ Group, labeller=label_both) +
  theme(legend.position = "none") +
  scale_y_continuous("standardized weight")

```


It is easier to compare how the rats gained weight compared to the other rats. In the plot before, it looked like the rats were steadily gaining weight. However, in the new plot, we can see that it was not true for all cases. Next, we can perform a t-test to see more clearly the difference between the groups. As we know, there is a big rat (outlier) in group two, which may disturb the results. However, if the rat is dropped, the sample size is reduced. Hence, no dropping will be done.

