

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# **Exercise 4. Clustering and classification**

The data used in this exercise comprises the housing values in the suburbs of Boston. The Boston data frame has 506 rows and 14 variables. 

```{r}
library(magrittr)
library(MASS)
library(ggplot2)
library(tidyr)
library(corrplot)
library(GGally)


data("Boston")
str(Boston)
summary(Boston)



#graphical overview
cor_matrix<-cor(Boston) %>% round(digits = 2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

From the summary we can see that the variables are numerical. The chas variable is a dummy variable. From the graph we can see that the variable distributions are skewed. However the variable rm 

The correlation matrix shows the associations between the variables. The bigger and more colorful the circle is, the stronger the correlation is between the variables. There appears to be a strong positive relationship between property taxes (tax) and radial highway accessibility (rad). The lower status of the population (istat) and median value of owner-occupied homes in $1000s (medv) have a strong negative association. Additionally, strong negative association is between proportion of owner-occupied units built prior to 1940 (age) and weighted mean of distances to five Boston employment centres (dis).

