---
title: "IODS course project"
author: "Ida Alakörkkö"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# *Introduction to Open Data Science - Course Project*

Introduction to Open Data Science is a course for anyone interested in learning how to use RStudio, R Markdown, and GitHub (state-of-the-art data science tools) to visualize and analyze open data using multivariate statistical methods while adhering to reproducible research principles and practices. 

I expect to master the fundamentals of Rstudio. I've never done any programming before.I am now getting a master's degree in health economics and working on my master's thesis. 

My GitHub repository: https://github.com/Idaala/IODS-project

```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***
```{r child = "chapter3.Rmd"}

```
***

```{r child= "chapter4.Rmd"}

```



(more chapters to be added similarly as we proceed with the course!)

